Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-26 ReviewerResponse.txt

In the following we respond specifically to the reviewers comments.

Reviewer 1:

As suggested, we made several changes to improve the structure of the paper.  
In summary, we have :
(1) added subsections in Section 3 [HSC: Can you say what these are?];
(2) gave an overview of the idea [HSC: idea? concept?] of the translator in Section 4;
(3) removed unnecessary Algorithm 1 in Section 5 [HSC: give further explanation here, why is it unnecessary?];
(4) added a subsection in Section 5 for a better outline [HSC: give name of the subsection.........]; and 
(6) further explained Algorithm 3 (now Algorithm 2) [HSC: more details, what did you further explain?].

We also gave further detailed descriptions about the PoM-Tagger, 
namely its parsing process and lexicon files.

We also added LaTeXML to the related work section and gave a brief 
explanation with cited research (Schubotz et al., 2018, now included 
in our references), in order to explain why we did not choose MathML 
as an intermediate data format for performing translations. In that 
citation, we described detailed testing that we performed using LaTeXML 
and other state of the art LaTeX to MathML conversion tools. In that
reference, we discovered several critical problems which could occur 
when translations would be performed using MathML as an intermediate
data format.

Note that we did not add further comparisons with LaTeXML to our 
evaluations section, because current CAS are unable to understand MathML 
which contains special functions. Thus we can only compare translations 
without function calls or implement a solution which overcomes this 
problem, in order to make a decent comparison. Because of the great amount 
of problems which would need to be overcome, a full solution translation 
implementation using MathML is out of the scope of our current project. 
This is now all explained in detail in the new version of the manuscript.

Reviewer 2:

For clarification, throughout the manuscript, we decided to change the 
wording "Word Processor" to "Document Preparation System." Since this wording
occurred as part of the title, we did not add that we focus on LaTeX to the 
title and hope that the Abstract points this out in an acceptable manner.

Also, instead of creating an Appendix, we decided to add short paragraphs 
which explain the purpose of the sections [HSC: What sections?]. We hope 
this helps the reader to decide whether he can skip the section, e.g., in 
case of his lack of interest in Maple "hackings", or not.  In the new 
version of the paper, we implemented all suggested minor changes as well as 
adding the idea of improved numerical tests by using the mentioned cut 
region detection to the future work section.

### 2018-11-26 main.tex
made some changes to main.tex

\begin{document}
\maketitle
\begin{abstract}
\glsresetall
\noindent
\textbf{Purpose:} Modern mathematicians and scientists of math-related disciplines use \glspl*{dps} to write and \glspl*{cas} to calculate mathematical expressions. Usually, they translate the expressions manually between \gls*{dps} and \gls*{cas}. This process is time-consuming and error-prone. Our goal is to automate this translation. This paper uses \Maple{} and \Mathematica{} as the \gls*{cas}, and \LaTeX{} as our \gls*{dps}.

\noindent\textbf{Approach:} The \gls*{nist} developed a collection of special \LaTeX{} macros that create links from mathematical symbols to their definitions in the \gls*{nist} \gls*{dlmf}. We are using these macros to perform rule-based translations between the formulae in the \gls*{dlmf} and \gls*{cas}. Moreover, we develop software to ease the creation of new rules and to discover inconsistencies.

@@

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use a \gls*{dps} to write the paper and one or more \glspl*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX{}
%\footnote{Note that technically \LaTeX{} is not a \gls*{dps} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{dps} as well.} 
has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in{\mathbf R}$.
This mathematical expression is written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
	\mathscr{W} \left\{ w_1(z), w_2(z) \right\} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation becomes unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

@@

The translation process of the \Macro 's is complex, so there is a special class, the \verb|MacroTranslator|, that handles those nodes in the \gls*{pom-pt}. Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has, in Line~\ref{line:get_info}. It also processes the following siblings to translate the arguments. The \verb|MacroTranslator| will be called in Line~\ref{line:transLeaf} in Algorithm~\ref{alg:translation}, since the macro is a leaf node in the \gls*{pom-pt}. The following cases describes the different kinds of the following siblings after a semantic macro node. Those can be
\begin{itemize}
\item an exponent, such as for `\verb|^2|' right after the macro node (Line~\ref{line:next_caret});
\item an optional parameter in square brackets right after the macro node or after an exponent (Line~\ref{line:next_optional});
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above and no `$@$' symbols were passed yet (Line~\ref{line:trans_paras});
\item `$@$' symbols (Line~\ref{line:skip_ats}); or
\item a variable in curly brackets (a \textit{sequence} node) after the `$@$' symbols were passed. (Line~\ref{line:trans_vars})
\end{itemize}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to get access to the parse tree of \Maple's input. The translation process from this parse tree than follows the same principle as for the forward translations.

@@

With this technique we can detect translation errors such as in \eqref{eq:wrong-trans-1} and \eqref{eq:wrong-trans-2}. Consider the \gls*{dlmf} equation for the sine and cosine function~\parencite[(4.21.2)]{NIST:DLMF}
\begin{equation}
\sin \left(u+v\right) = \sin{u}\cos{v} + \cos{u}\sin{v}.
\end{equation}
Assume the translator would forward translate the expression based on (\ref{eq:wrong-trans-1},~\ref{eq:wrong-trans-2}). Than
\begin{eqnarray}
\verb|\sin@{u + v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u + v)|,\\
\verb|\sin@@{u}\cos@@{v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u)*sin(v)|,\\
\verb|\cos@@{u}\sin@@{v}| & \overset{\langMaple}{\mapsto} & \verb|sin(u)*cos(v)|.
\end{eqnarray}
This produces the equation in \Maple
\begin{equation}
\cos\left(u+v\right) = \cos{u}\sin{v} + \sin{u}\cos{v},
\end{equation}
which is wrong. Since the expression is correct before the translation, we conclude an error during the translation process.

>>>>

\begin{document}
\maketitle
\begin{abstract}
\glsresetall
\noindent
\textbf{Purpose:} Modern mathematicians and scientists of math-related disciplines often use \glspl*{dps} to write and \glspl*{cas} to calculate mathematical expressions. Usually, they translate the expressions manually between \gls*{dps} and \gls*{cas}. This process is time-consuming and error-prone. Our goal is to automate this translation. This paper uses \Maple{} and \Mathematica{} as the \gls*{cas}, and \LaTeX{} as our \gls*{dps}.

\noindent\textbf{Approach:} Bruce Miller at \gls*{nist} developed a collection of special \LaTeX{} macros that create links from mathematical symbols to their definitions in the \gls*{nist} \gls*{dlmf}. We are using these macros to perform rule-based translations between the formulae in the \gls*{dlmf} and \gls*{cas}. Moreover, we develop software to ease the creation of new rules and to discover inconsistencies.

@@

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use a \gls*{dps} to write the paper and one or more \glspl*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX{}
%\footnote{Note that technically \LaTeX{} is not a \gls*{dps} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{dps} as well.} 
has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in{\mathbf R}$.
This mathematical expression can be written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
\mathscr{W}\{ w_1(z), w_2(z) \} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation becomes unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

@@

The translation process of the \Macro 's is complex, so there is a special class, the \verb|MacroTranslator|, that handles those nodes in the \gls*{pom-pt}. Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has, in Line~\ref{line:get_info}. It also processes the following siblings to translate the arguments. The \verb|MacroTranslator| will be called in Line~\ref{line:transLeaf} in Algorithm~\ref{alg:translation}, since the macro is a leaf node in the \gls*{pom-pt}. The following cases describes the different kinds of the following siblings after a semantic macro node. Those can be
\begin{itemize}
\item an exponent, such as for `\verb|^2|' right after the macro node (Line~\ref{line:next_caret});
\item an optional parameter in square brackets right after the macro node or after an exponent (Line~\ref{line:next_optional});
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above and no `$@$' symbols were passed yet (Line~\ref{line:trans_paras});
\item `$@$' symbols (Line~\ref{line:skip_ats}); or
\item a variable in curly brackets (a \textit{sequence} node) after the `$@$' symbols were 
passed (Line~\ref{line:trans_vars}).
\end{itemize}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to get access to the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations.

@@

With this technique we can detect translation errors such as in \eqref{eq:wrong-trans-1} and \eqref{eq:wrong-trans-2}. Consider the \gls*{dlmf} equation for the sine and cosine function~\parencite[(4.21.2)]{NIST:DLMF}
\begin{equation}
\sin \left(u+v\right) = \sin{u}\cos{v} + \cos{u}\sin{v}.
\end{equation}
Assume the translator would forward translate the expression based on (\ref{eq:wrong-trans-1},~\ref{eq:wrong-trans-2}). Then
\begin{eqnarray}
\verb|\sin@{u + v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u + v)|,\\
\verb|\sin@@{u}\cos@@{v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u)*sin(v)|,\\
\verb|\cos@@{u}\sin@@{v}| & \overset{\langMaple}{\mapsto} & \verb|sin(u)*cos(v)|.
\end{eqnarray}
This produces the equation in \Maple
\begin{equation}
\cos\left(u+v\right) = \cos{u}\sin{v} + \sin{u}\cos{v},
\end{equation}
which is wrong. Since the expression is correct before the translation, we conclude an error during the translation process.
